---
title: "Structural equation modelling"
author: Valerii Zuev
date: 11/28/2020
output: html_notebook
---
Preparing workspace:
```{r warning = F, message = F}
library(here) # for relative paths
library(tidyverse) # dplyr, ggplot2, ...
library(lavaan) # confirmary factor analysis
library(ggcorrplot) # special extension methods for ggplot2

source(here("r/biomod-prac/task6_sem/ggsem.R"))
df <- here("r/biomod-prac/task4-2020-nov-01/nut.csv") %>%
  read.csv(header = T) %>%
  select(PodWidth, PodLength, SeedsWeight, SeedsNumber, PodsNumber, PodsWeight, BegFEndF, EndFBegM) %>%
  na.omit

df
```

Finding dependent pairs of variables by their correlations:

```{r}
df %>% cor %>% ggcorrplot
```
We see that all variables form three clusters, within each of which variables are strongly correlated:

1. PodLength, PodWidth
2. PodsWeight, PodsNumber, SeedsNumber, SeedsWeight
3. EndFBegM, BegFEndF


Building a model of equations:
```{r}
HS.model <- " PodsDimension =~ PodLength + PodWidth
              NumbersAndWeights =~ PodsWeight + PodsNumber + SeedsNumber + SeedsWeight
              EndF =~ EndFBegM + BegFEndF"

fit <- cfa(HS.model, data = df)
summary(fit, fit.measures = TRUE)
```

RMSEA = 0.290 > 0.06, which is not perfect.

The chi-squared test p-value is equal to zero (accurate to 3 decimal places).

All this indicates that the model should not be accepted. Nevertheless, we will plot the results:

```{r message = F, warning = F, r, fig.width=10, fig.height=7}
ggsem(fit) +
  ggtitle("Nut dataset: structural equation model
  (for selected features)") +
  theme(plot.title = element_text(hjust = 0.5))
```
