---
title: "Principal Component Analysis"
author: Valerii Zuev
date: 11/19/2020
output: html_notebook
---

Preparing workspace:

```{r message = F, warning = F}
library(here) # for relative paths
library(types) # type hints
library(tidyverse) # dplyr, ggplot2, ...
library(pca3d) # Principal Component Analysis
library(factoextra) # for `fviz_contrib`

df <- here("r/biomod-prac/task4-2020-nov-01/nut.csv") %>%
  read.csv(header = T) %>%
  na.omit
```

Computing principal components using a library method:

```{r}
lib.pca <- df %>% prcomp(center = TRUE, scale = TRUE)
lib.pca_data <- data.frame(var = lib.pca$sdev^2) # variances
```

Custom implementation of PCA (using a built-in singular value decomposition):
```{r}
df_scaled <- df %>% scale %>% t %>% as.data.frame
s <- svd(df_scaled)
my.pca_x <- s$v # principal components
my.pca_data <- data.frame(var = s$d^2 / length(df_scaled)) # variance
```

*Explanation*. Let $X \in \mathbb{R}^{n\times m}$ denote our data (columns = features, rows = samples) <br>
Computing a *singular value decomposition* of $X$:
$$X = U \times D \times V^T$$
where
$$U: U^T U=1, \hspace{5pt} U \in \mathbb{R}^{n\times m} \\
V: V^T = V^{-1}, \hspace{5pt} V \in \mathbb{R}^{m \times m} \\
D = diag(\sigma_1, ..., \sigma_m)$$

Then $V$ is a matrix of principal components and $D^2 /n$ is a matrix which main diagonal contains new variances for each component.

Reducing dimensions to leave only 75% of variances:

```{r}
n_main_components <- function(variances = ? numeric, keep_variance = .75) {
  total_var <- sum(variances)
  retained_var <- 0
  for (i in seq_along(variances)) {
    retained_var <- retained_var + variances[i]
    if (retained_var / total_var >= keep_variance) {
      return(i)
    }
  }
  return(-1) # error, could not keep the specified amount of variance
}

my.n_comp <- n_main_components(my.pca_data$var)
lib.n_comp <- n_main_components(lib.pca_data$var)
cat(paste("number of components to keep: ", my.n_comp, "(custom implementation), ", lib.n_comp, "(library implementation)"))
my.main_components <- my.pca_x[, 1:my.n_comp] %>% t %>% as.data.frame
lib.main_components <- lib.pca$x[, 1:lib.n_comp] %>% t %>% as.data.frame
```

Plotting the explained variation:
```{r message = F, warning = F}
plot_explained_var <- function(pca_data = ? data.frame, title = ? character) {
  sum_var <- sum(pca_data$var)
  ggplot(data = pca_data, aes(x = seq_along(var), y = cumsum(var) / sum_var)) +
    geom_line() +
    geom_point() +
    geom_hline(yintercept = .75, colour = "red") +
    xlab("number of components") +
    ylab("retained variance") +
    ggtitle(paste0("Proportion of explained variation (acceptable threshold marked with red)\n", title)) +
    theme(plot.title = element_text(hjust = 0.5))
}

plot_explained_var(lib.pca_data, "(library implementation)")
plot_explained_var(my.pca_data, "(custom implementation)")
```
Which feature has the most effect on the first component?
```{r}
fviz_contrib(lib.pca, choice = "var", axes = 1) +
  labs(title = "Contributions to Dim 1")
```
We see that it is `NoPodsWeight`

Plotting data on the plane of two first principal components:

```{r}
most_contrib_idx <- match("NoPodsWeight", colnames(df))

plot_two_main <- function(main_components = ? data.frame, title = ? character) {
  two_comp <- data.frame(PC1 = main_components[, 1],
                         PC2 = main_components[, 2],
                         color = integer(nrow(main_components)),
                         size = integer(nrow(main_components)) + 1
  )
  two_comp$color[most_contrib_idx] <- 1
  two_comp$size[most_contrib_idx] <- 2
  ggplot(data = two_comp, aes(x = PC1, y = PC2, col = color, size = size)) +
    geom_point() +
    xlab("principal component #1") +
    ylab("principal component #2") +
    ggtitle(paste0("Two first principal components
    (blue point - feature with the highest contribution to pr. comp. 1)\n", title)) +
    theme(plot.title = element_text(hjust = 0.5), legend.position = "none")
}
plot_two_main(lib.main_components, "(library implementation)")
plot_two_main(my.main_components, "(custom implementation)")
```

