---
title: "RNA-Seq part II / Differential expression analysis"
author: Valerii Zuev
date: 4/19/2021
output:
    html_notebook:
        toc: true
        includes:
            in_header: header_favicon.html
---

## Introduction

This script works with data obtained from the 2016 study "Candidate gene networks and blood biomarkers of
methamphetamine-associated psychosis: an integrative
RNA-sequencing report" ([doi:10.1038/tp.2016.67](https://doi.org/10.1038/tp.2016.67)).

## System requirements

The code was written and tested on a Windows 10 machine using R version 4.0.2 (2020-06-22) for x86_64-w64-mingw32 platform with BiocManager version 3.12.

Substitute the path in `abs_path` function with an appropriate path in your environment.

## Setup

Clone the repository from GitHub:

```
git clone https://github.com/zuevval/source.git
```
Then navigate to `r/bioinf2021/lab2rna_seq` directory which contains the source code for this script.

The following files are necessary to reproduce our results:

* `annot.csv` (already present in the `dataset` subdirectory)
* `GSE74737_counts.txt`. This table may be retrieved from the first page of the .xls file available at [GEO site](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE74737) or directly downloaded from [azuev.ddns.net/~valera/stu/2021bio/rnaseq](http://azuev.ddns.net/~valera/stu/2021bio/rnaseq/)
* `c5.go.v7.4.entrez.gmt` (all GO gene sets as NCBI (Entrez) gene IDs) and `c2.cp.kegg.v7.4.entrez.gmt` (KEGG gene sets as NCBI (Entrez) gene IDs).
May be obtained from [gsea-msigdb.org](http://www.gsea-msigdb.org/gsea/downloads.jsp) or from [azuev.ddns.net/~valera/stu/2021bio/rnaseq](http://azuev.ddns.net/~valera/stu/2021bio/rnaseq/)

You will need R packages listed below:

* tidyverse (`install.packages("tidyverse")`)
* here (`install.packages("here")`)
* DESeq2 (`BiocManager::install("DESeq2")`)
* AnnotationDbi (`BiocManager::install("AnnotationDbi")`)
* org.Hs.eg.db (`BiocManager::install("org.Hs.eg.db")`)
* fgsea (`BiocManager::install("fgsea")`)

## Analysis

### Preparing workspace

first, we define the directory which contains `annot.csv` and `GSE74737_counts.txt`
```{r message=F, warning=F, output=F}
library(tidyverse)
abs_path <- function(path) paste0("E:/Users/vzuev/github-zuevval/source/r/bioinf2021/lab2rna_seq/dataset/",path)
```
Then we load these files into `data.frame` objects (and filter columns in Counts leaving only those present in `annot.csv` rows)

```{r}
coldata <- read.csv(abs_path("annot.csv"), row.names = 1) %>% mutate_all(as.factor)
head(coldata)
counts_df <- read.csv(abs_path("GSE74737_counts.txt"), sep = "\t", row.names = "Gene") %>%
  select(rownames(coldata))
head(counts_df)
```
### Processing RNA-seq data with DESeq2

Genes in `GSE74737_counts.txt` are identified by HUGO gene symbols.
We convert them to Entrez IDs (here's where you need `AnnotationDbi`)
```{r message=F, warning=F, output=F}
counts <- counts_df %>%
  mutate(entrez = AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, rownames(counts_df), "ENTREZID", "SYMBOL")) %>%
  filter(!is.na(entrez)) %>%
  select(-entrez) %>%
  as.matrix
```
```{r}
head(as.data.frame(counts))
```
Performing a sanity check:
```{r}
assertthat::assert_that(all(rownames(coldata) == colnames(counts)))
```
Finally, we construct a `DESEqDataSet` object and feed it to `DESeq2::DESeq(...)` which perform the analysis.
Since there is a single factor (condition), we use it as the only parameter for the model design.
```{r message = F, warning = F}
dds <- DESeq2::DESeqDataSetFromMatrix(countData = counts, colData = coldata, design = ~condition)
dds_analysed <- DESeq2::DESeq(dds)
deseq_res <- DESeq2::results(dds_analysed)
deseq_res_df <- as.data.frame(deseq_res)
head(deseq_res_df %>% arrange(padj))
```
None of the genes show an adjusted p-value lower than the threshold specified in the article (p-value < 0.01) ¯\\\_(ツ)\_/¯

so we'll select and save those that are less than the 0.2 threshold:

```{r message = F, warning = F, output = F}
pval_thresh <- .2
deseq_res_df$entrez <- AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, rownames(deseq_res_df), "ENTREZID", "SYMBOL")
diff_expr <- deseq_res_df %>% filter(padj < pval_thresh)
devnull <- mapply(function(df = ? data.frame, filename = ? character) {
  df %>%
    select(entrez) %>%
    unlist %>%
    write.table(abs_path(paste0(filename, "_entrez.csv")), sep = ",", row.names = F, col.names = F, quote = F)
  df %>%
    rownames %>%
    unlist %>%
    write.table(abs_path(paste0(filename, "_symbols.csv")), sep = ",", row.names = F, col.names = F, quote = F)
}, list(deseq_res_df, diff_expr), c("background", "diff_expr"))

```
### GO and KEGG enrichment analysis

Loading the abovementioned pathways:

```{r}
parse_gmt <- function(gmt_filename = ? character){
  strlist <- strsplit(readLines(abs_path(gmt_filename)), "[[:space:]]+")
  pathways <- lapply(strlist, tail, n = -2)
  names(pathways) <- lapply(strlist, head, n = 1)
  pathways
}
kegg_pathways <- parse_gmt("c2.cp.kegg.v7.4.entrez.gmt")
go_pathways <- parse_gmt("c5.go.v7.4.entrez.gmt")
```

Creating ranks:

```{r}
res_sorted <- deseq_res_df %>% arrange(padj)
ranks <- res_sorted$padj
names(ranks) <- res_sorted$entrez
save(ranks, file=abs_path("ranks.RData"))
```

Running FGSEA:

```{r message = F, warning = F}
load(abs_path("ranks.RData"))
fgsea_kegg_res <- fgsea::fgsea(kegg_pathways, ranks, scoreType = "pos")
fgsea_kegg_res %>% arrange(padj) %>% head

fgsea_go_res <- fgsea::fgsea(go_pathways, ranks, scoreType = "pos")
fgsea_go_res %>% arrange(padj) %>% head
```

Saving filtered pathways (associated with our gene set with adjusted p-value < 0.05):

```{r}
padj_pathways_thresh <- .05
devnull <- mapply(function(df = ? data.frame, filename = ? character) {
  df %>%
    select(pathway, padj) %>%
    arrange(padj) %>%
    filter(padj < padj_pathways_thresh) %>%
    write.table(abs_path(paste0(filename, "_pathways.csv")), sep = ",", row.names = F, quote = F)
}, list(fgsea_go_res, fgsea_kegg_res), c("go", "kegg"))

```
